Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-08-30 aaai-2018/khodadadi-mcdonald-aaai19.tex
initial aaai version

\documentclass{article}
\PassOptionsToPackage{sort,numbers,square}{natbib}

\usepackage{nips_2018}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{hyperref}
\hypersetup{backref,colorlinks=true,citecolor=blue,linkcolor=blue,urlcolor=blue}

%\newcommand{\theHalgorithm}{\arabic{algorithm}}

\usepackage{graphicx}
\usepackage{subfigure} 
\graphicspath{{../nips-2018/}}

@@

\renewcommand*{\figureautorefname}{Figure}%
\renewcommand*{\tableautorefname}{Table}%
\renewcommand*{\sectionautorefname}{Section}%
\renewcommand*{\subsectionautorefname}{Section}%
\renewcommand*{\subsubsectionautorefname}{Section}%

@@

@@

\usepackage{xr}
\externaldocument{paper-supplement}

\title{Modeling trend in temperature volatility using generalized LASSO}

\author{Arash Khodadadi\\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{arakhoda@indiana.edu}
\And
  Daniel J.\ McDonald \\
 Department of Statistics\\
 Indiana University\\
 Bloomington, IN 47408 \\
 \texttt{dajmcdon@indiana.edu}}

@@

@@

\begin{abstract}
In this paper, we present methodology for estimating trends in
spatio-temporal volatility. We give two algorithms for computing our
estimator which are tailored for dense, gridded observations over both
space and time, though these can be easily extended to other
structures (time-varying network flows, neuroimaging). We motivate our
methodology by applying it to a massive climate dataset and
discuss the implications for climate analysis.
\end{abstract}

@@

Estimating smooth trends over time for large collections of time
series is a common problem in economics, finance, meteorology,
neuroscience and more. Most of this work focuses on analyzing trends
(or removing trends) in the temporal average, but trends in variance
can be more relevant, especially for financial data but also for
climate science.

Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature~\citep{huntingford_no_2013},
because, an increase in the
temperature variability will increase the probability of extreme hot
outliers~\citep{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\citep{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, increased frequency of extreme cold events in the
northern hemisphere can be partially attributed to increases in
mean climate but is also due to increases in temperature
variance~\citep{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citep{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set by
examining the empirical SD at each spatial location relative to that
location's SD over a base period, and showed
that these estimates are increasing.
\citep{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. Their results showed that there
still is an increase in the SDs from 1958-1970 to 1991-2001, but this
is much less than what is obtained from the method used in
\citep{hansen_perception_2012}. The authors also computed the
time-evolving global SD from the de-trended time-series at each
position, which suggests that the global SD has been
stable.

These and other related research, e.g.,
\citep{rhines_frequent_2013}, have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between the
observations are ignored.

\subsection{Related work}

Variance estimation for financial time series has a lengthy history,
focused especially on parametric models like the generalized
autoregressive conditional heteroskedasticity (GARCH) process~\citep{engle2002dynamic} and
stochastic volatility models~\citep{HarveyRuiz1994}. These models (and
related AR processes) are specifically for parametric modelling of
short ``bursts'' of high volatility, behavior typical of financial
instruments. Parametric models for spatial data go back at least
to~\citep{besag1974spatial} who proposed a conditional probability
model on the lattice for examining plant ecology.

More recently, nonparametric models for both spatial and temporal data
have focused on using $\ell_1$-regularization for trend
estimation. \citep{KimKoh2009} proposed $\ell_1$-trend filtering for
univariate time series, which forms the basis of our methods. These
methods have been generalized to higher order temporal smoothness
~\citep{Tibshirani2014}, graph dependencies~\citep{WangSharpnack2016},
and, most recently, small, time-varying
graphs~\citep{HallacPark2017}. Our methodology is similar in flavor
to~\citep{HallacPark2017}, though it uses a different likelihood
function to emphasize variance estimation rather than trends in mean
signal. Furthermore, the focus is in high-dimensional, regular data
rather than a small number of changing graphs.

\subsection{Main contributions}

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time is considered as
a hidden variable. The values of these hidden variables
are then estimated by maximizing the likelihood of the observed
data. We show that this formulation is not appropriate for
detecting the trend, so, following ~\citep{Tibshirani2014}, we penalize the
differences between the estimated variances which
are temporally and spatially ``close'', resulting
in a generalized LASSO problem. % However, the
% dimension of this optimization problem is very high and so the
% standard methods solvers are inadequate. We develop two algorithms
% which are computationally feasible. In the first method, we adopt an optimization technique
% called alternative direction method of multipliers (ADMM)
% \citep{boyd_distributed_2011}, to divide the total problem into
% several sub-problems of much lower dimension and show how the total
% problem can be solved by iteratively solving these sub-problems. The
% second method, called the \textit{linearized ADMM algorithm}
% \citep{parikh_proximal_2014} solves the main problem by iteratively
% solving a linearized version of it. We will compare the benefits of
% each method.

Our main contributions are as follows:
\begin{enumerate}
\item We propose a model for nonparametric variance estimation for a
  spatio-temporal process (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms (ADMM) to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful.
\item We illustrate our methods on a large global temperature dataset
  with the goal of tracking world-wide trends in variance as well as a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

While we motivate and illustrate our methods on large, gridded climate
data, we note that our algorithms are also applicable to neuroimaging
data or large collections of financial instruments.

\section{Estimating the variance of spatio-temporal data}
\label{sec:ell_1-trend-filt}

$\ell_1$-trend filtering was proposed by \citep{KimKoh2009} as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D\beta}_1
\label{eq:l1tf}
\end{equation}
 where $y_t$ is an observed time-series, $\beta$ is the smooth trend,
 $D$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise
linear. \citep{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} is a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on $\beta$.

@@

\subsection{Estimating the variance}
\label{sec:l1tf_var}

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a hidden
variable $h_t$ such that conditioned on $h_t$ the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the hidden variables $h_t$ vary smoothly. To impose
this assumption, we estimate $h_t$ by solving the penalized, negative
log-likelihood: 
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ Dh }_1
\label{eq:l1tf_var}
\end{equation}
 where $D$ has the same structure as above.

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. Due to space limitations, we
defer details to Appendix A in the Supplement, where we show how to derive the dual of this optimization problem and compute the first and second derivatives of the dual objective function.

@@

\subsection{Adding spatial constraints}
\label{sec:exten}

@@

At a specific time $t$, the data is measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
\begin{align}
\min_h &\sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}}
+\lambda_1 \sum_{i,j} \sum_{t=1}^{T-2} \left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\
&\quad+\lambda_2 \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_2 \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\label{eq:l1tf_var_st}.
\end{align}

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on graphs~\citep{WangSharpnack2016} which penalizes the difference between the estimated values of the signal on the connected nodes. As before, we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times S$ vector and $D$ is replaced by $D_{ST} \in \mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and spatial constraints, respectively\footnote{$N_s$ is obtained by counting the number of unique constraints at each location and at all times.}. The exact form of this matrix in clarified in Appendix A in the Supplement. Then, as we have two different tuning parameters for the temporal and spatial components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;  \lambda_2\one_{N_s}^\top\right]^\top$ leading to:\footnote{Throughout the paper, we use $|x|$ for both scalars and vectors. For vectors we use this to denote a vector obtained by taking the absolute value of each entry of $x$.} 
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{ST}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}

\section{Proposed optimization methods}
\label{sec:prop-optim-meth}

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $n_r=90$, $n_c=360$, $T=3650$ and so $D_{ST}$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations which
depends on $D_{ST}^\top D_{ST}$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that this is a highly structured and sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

In the next section, we develop two ADMM algorithms for solving this
problem efficiently. The first casts the problem as a
so-called consensus optimization problem~\citep{boyd_distributed_2011}
which solves smaller sub-problems using PDIP and then recombines the results. The
second uses proximal methods to avoid matrix inversions.

\subsection{Consensus optimization}
\label{sec:consOpt}

\begin{comment}
General (unconstrained) optimization problems $\min_z
\sum_i f_i(z)$, $z\in\mathbb{R}^n$ can be rewritten equivalently using a
collection of local variables, $x_i\in \mathbb{R}^{n_i}$, and a
constraint. Following the notation of 
\citep{boyd_distributed_2011}, let $k=\mathscr{G}(i,j)$ which
means that the $j^\text{th}$ entry of $x_i$ is $z_k$ (or
$(x_i)_j=z_k$) and define $\tilde{z}_i \in \mathbb{R}^{n_i}$ by
$(\tilde{z}_i)_j=(x_i)_j$. Then the original unconstrained
optimization problem is equivalent to the following constrained
problem
$\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)$ subject to $\tilde{z}_i=x_i$.
Now, we can apply ADMM to the augmented Lagrangian of this
problem: 
\begin{align}
x_i&\leftarrow \argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
     (\rho/2) \norm{ x_i-\tilde{z}_i }_2^2\\
%\label{eq:ADMM_steps}
z_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j\\
u_i&\leftarrow u_i + \rho (x_i-\tilde{z}_i)
\end{align}
\end{comment}

>>>>

%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (Algorithms for Estimating Trends in Global Temperature Volatility)
%/Author (Khodadadi and McDonald)
/Author (John Doe, Jane Doe)}
\setcounter{secnumdepth}{2}

% \usepackage[bookmarks=false]{hyperref}       
% \AtBeginDocument{\NoHyper} % turn off all links, colors
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

% For citations
%\usepackage[sort,numbers,square]{natbib}

\newcommand{\citet}[1]
{\citeauthor{#1} ̃\shortcite{#1}}
\newcommand{\citep}{\cite}
\newcommand{\citealp}[1]
{\citeauthor{#1} ̃\citeyear{#1}}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
%\hypersetup{backref,colorlinks=true,citecolor=blue,linkcolor=blue,urlcolor=blue}

%\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{cleveref}

\usepackage{subfigure} 
\graphicspath{{../Figures/}}

@@

@@

\newcommand{\autoref}{\Cref}

@@

\usepackage{xr}
\externaldocument{aaai-paper-supplement}

@@

\title{Algorithms for Estimating Trends in Global Temperature Volatility}

\author{Author 1 \ and Author 2\\
Address line\\
Address line
}
% \author{Arash Khodadadi \and Daniel J. McDonald\\
%  Department of Statistics\\
%  Indiana University\\
%  Bloomington, IN 47408 \\
%  \{arakhoda,dajmcdon\}@indiana.edu}

@@

\begin{abstract}
Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature. In this paper, we develop methodology for estimating such
trends using multi-resolution climate data from polar orbiting weather
satellites. We derive two novel algorithms for computation that are
tailored for dense, gridded observations over both 
space and time. We evaluate our methods with a simulation constructed
to mimic these data's 
features and on a large, publicly available, global
temperature dataset with the goal of tracking trends in
cloud reflectance temperature variability.
\end{abstract}

@@

% Estimating smooth trends over time for large collections of time
% series is a common problem in economics, finance, meteorology,
% neuroscience and more. Most of this work focuses on analyzing trends
% (or removing trends) in the temporal average, but trends in variance
% can be more relevant, especially for financial data but also for
% climate science.

The amount of sunlight reflected from clouds is among the largest
sources of uncertainty in climate
prediction~\citep{BoucherRandall2013}. But climate models fail to
reproduce global cloud statistics and understanding the reasons for
this failure is a grand challenge of the World Climate Research
Programme~\citep{BonyStevens2015}.
By understanding how cloudliness changes over time, examining changes
in brightness temperature within various cloud types, and inferring
changes in cloud radiative effects, atmospheric scientists can better model climate
change. Modeling inter-satellite biases, satellite drift, and
seasonal and long-term climate phenomena like El Ni\~{n}o will lead to
a better understanding of climate
change~\citep{SchreierKahn2014,BaumWielicki1994,BaumWielicki1992,FreyAckerman1996}.

Numerous
studies have examined the overall impacts of clouds on climate
variability~\citep{MyersMechoso2018,GrisePolvani2013,BenderRamanathan2012},
but such investigations have been hampered by the lack of a suitable
dataset. Ideal data would have global coverage at high spatial
resolution, a long enough record to recover temporal trends, and be
multispectral~\citep{WielickiYoung2013}. The International Satellite
Cloud Climatology Project (ISCCP)
has been producing cloud property information from
limited spectral channels for over three
decades~\citep{RossowSchiffer1991}. These data have been used in 
various climate variability
studies~\citep[e.g.]{BenderRamanathan2012}, however, the utility of
these data for long-term study of climate variability has been
questioned~\citep{EvanHeidinger2007}.

ISCCP provides a relatively long record, but it only incorporates
radiances from limited visible and infrared channels from Advanced
Very High Resolution Radiometer (AVHRR) instruments. Ongoing work seeks to create a
more spectrally-detailed dataset which can avoid the above issues by
combining radiance data from AVHRR imagers with readings from High-resolution
Infrared Radiation Sounders on board legacy
satellites~\citep{StatenKahn2016,SchreierKahn2010,KahnFishbein2007}. In
anticipation of this new dataset, our work develops novel methodology
for examining the trends in variability of climate data across space
and time.

\subsection{Variability Rather Than Average}
\label{sec:vari-rath-than}

Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature~\cite{huntingford_no_2013},
because an increase in
temperature variability will increase the probability of extreme hot
outliers~\cite{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\cite{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, a point of active debate is the extent to which the observed
increased frequency of extreme cold events in the 
northern hemisphere can be attributed to increases in temperature
variance  rather than to increases in
mean climate~\cite{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citet{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set by
examining the empirical SD at each spatial location relative to that
location's SD over a base period, and showed
that these estimates are increasing.
\citet{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. Their results showed that there
still is an increase in the SDs from 1958-1970 to 1991-2001, but this
is much less than what is obtained from the method used in
\cite{hansen_perception_2012}. The authors also computed the
time-evolving global SD from the de-trended time-series at each
position, which suggests that the global SD has been
stable.

These and other related research, e.g.,
\citet{rhines_frequent_2013}, have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between the
observations are completely ignored.

In the present work, we examine variance (rather than the mean) for a number of reasons.
First, instrument bias in the satellites increases over time so
examining the mean over time conflates that bias with any actual
change in mean (though the variance is unaffected). 
Second, extreme weather events (hurricanes, droughts, wildfires in
California, heatwaves in Europe) may be driven more strongly by increases
in variance than by increases in mean. 
Finally, even if the global mean temperature is constant, there may still
be climate change. In fact, atmospheric physics suggests that,
across space, average temperatures should not change (extreme cold
in one location is offset by heat in another). But if swings across
space are becoming more rapid, then, even with no change in mean global
temperature over time, increasing variance can lead to
increases in the prevalence of extreme events.

\subsection{Main Contributions}

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time is considered as
a hidden variable. The values of these hidden variables
are then estimated by maximizing the likelihood of the observed
data. Following~\cite{Tibshirani2014}, we penalize the
differences between the estimated variances which
are temporally and spatially ``close'', resulting
in a generalized LASSO problem. However, in our application, the
dimension of this optimization problem is massive, and so the
standard solvers are inadequate. We develop two algorithms
which are computationally feasible. In the first method, we adopt an optimization technique
called alternating direction method of multipliers
(ADMM, \citealp{boyd_distributed_2011}), to divide the total problem into
several sub-problems of much lower dimension and show how the total
problem can be solved by iteratively solving these sub-problems. The
second method, called the \textit{linearized ADMM algorithm}
\cite{parikh_proximal_2014} solves the main problem by iteratively
solving a linearized version of it. We will compare the benefits of
each method.

Our main contributions are as follows:
\begin{enumerate}
\item We propose a model for nonparametric variance estimation for a
  spatio-temporal process and discuss the relationship between our
  methods and those existing in the machine learning literature (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms (ADMM) to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful. Open-source
  Python code is available on {\tt github}.
\item We illustrate our methods on a large, publicly available, global temperature dataset
  with the goal of tracking world-wide trends in variance as well as a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

While the motivation for our methodology is its application to large, gridded climate
data, we note that our algorithms are also applicable to neuroimaging,
image denoising, or examining large collections of financial instruments.

\section{Smooth Spatio-temporal Variance Estimation}
\label{sec:ell_1-trend-filt}

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D\beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise
linear. \citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

\subsection{Modifications for Variance}
\label{sec:l1tf_var}

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly. To impose
this assumption, we estimate $h_t$ by solving the penalized, negative
log-likelihood: 
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ Dh }_1
\label{eq:l1tf_var}
\end{equation}
where $D$ has the same structure as above.

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}, \citealp{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. Due to space limitations, we
defer details to Appendix A in the Supplement, where we show how to
derive the dual of this optimization problem and compute the first and
second derivatives of the dual objective function.

@@

\subsection{Adding Spatial Constraints}
\label{sec:exten}

@@

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\scriptsize
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}} +\lambda_1
\sum_{i,j} \sum_{t=1}^{T-2}
\left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\ 
&+\lambda_2 \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_2 \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is an
$T\times S$ vector and $D$ is replaced by $D_{ST} \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively\footnote{$N_s$ is obtained by
  counting the number of unique constraints at each location and at
  all times.}. The exact form of this matrix in clarified in Appendix
A in the Supplement. Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{ST}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}

\subsection{Related Work}

\attn{I've moved this here rather than above to focus the intro more
  on the data.}

Variance estimation for financial time series has a lengthy history,
focused especially on parametric models like the generalized
autoregressive conditional heteroskedasticity (GARCH) process~\cite{engle2002dynamic} and
stochastic volatility models~\cite{HarveyRuiz1994}. These models (and
related AR processes) are specifically for parametric modelling of
short ``bursts'' of high volatility, behavior typical of financial
instruments. Parametric models for spatial data go back at least
to~\cite{besag1974spatial} who proposed a conditional probability
model on the lattice for examining plant ecology.

More recently, nonparametric models for both spatial and temporal data
have focused on using $\ell_1$-regularization for trend
estimation. \citet{KimKoh2009} proposed $\ell_1$-trend filtering for
univariate time series, which forms the basis of our methods. These
methods have been generalized to higher order temporal smoothness
~\cite{Tibshirani2014}, graph dependencies~\cite{WangSharpnack2016},
and, most recently, small, time-varying
graphs~\cite{HallacPark2017}. Our methodology is similar in flavor
to~\cite{HallacPark2017}, though it uses a different likelihood
function to emphasize variance estimation rather than trends in mean
signal. Furthermore, the focus is in high-dimensional, regular data
rather than a small number of changing graphs.

\attn{Paragraph discussing relationship with estimating time-varying covariance}

\section{Optimization Methods}
\label{sec:prop-optim-meth}

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $S=90\times 360\approx 32,000$ spatial locations and
$T=3650$ time points, and so $D_{ST}$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations which
depends on $D_{ST}^\top D_{ST}$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that $D_{ST}$ is a highly structured, sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

In the next section, we develop two ADMM algorithms for solving this
problem efficiently. The first casts the problem as a
so-called consensus optimization problem~\cite{boyd_distributed_2011}
which solves smaller sub-problems using PDIP and then recombines the results. The
second uses proximal methods to avoid matrix inversions.

\subsection{Consensus Optimization}
\label{sec:consOpt}
